---
title: 'Homework #2'
author: "Francisco Brady"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
```

## Exercises

### 2.10
### 2.13
### 2.14
### 2.17
### 2.18
### 2.27
### 2.28
### 2.31

## Computing Question

```{r}
data <- read_csv('hw1_data.csv')
```

#### Plot each variable using a histogram and a boxplot. Describe the distributions of var1 and var2.

```{r}
var1hist <- ggplot(data) + geom_histogram(aes(var1))
var2hist <- ggplot(data) + geom_histogram(aes(var2))
var1hist / var2hist
```

Var 1 is bimodal in distribution. The values range from about 50 to around 140.  
Var 2 has a more unimodal distribution. The range is from -30 to a little bit over 20.  

```{r}
var1box <- ggplot(data) + geom_boxplot(aes(var1))
var2box <- ggplot(data) + geom_boxplot(aes(var2))
var1box / var2box
```

Var 1 has an IQR range of 120 to around 70 (50 units). The median of var 1 is a little over 95.
Var 2 has a smaller IQR and a tighter distribution. The median of var 2 looks to be around -1.


#### What type of plot do you think best illustrates each distribution? Why?  

The histogram is more useful for showing the distribution of the data in var 1. Relying only on the boxplot would obscure this about var 1. Because var 2 has a more unitary distribution and longer tails, the boxplot is appropriate to illustrate the spread of the data.  

#### Compute descriptive statistics for each variable. Use what you know about the distributions of the variables to decide which descriptive statistics are most appropriate.

```{r}
bind_rows(summary(data$var1),
          summary(data$var2)
) %>% knitr::kable(row.names = TRUE)

```

Because of the large range for both variables, it makes sense to report the median, which is less affected by large outliers, to get a sense of the central tendency of the values. It would also be useful to report the minimum and maximum of each variable to show the range of the data.  

